Dark matter halo shape at $z=0$ in the Auriga simulations: radial evolution and alignment with the stellar disk

### 2019-07-19 Document/paper.tex
edit
paper

% Abstract of the paper
\begin{abstract}
We present shape measurements of dark matter halos at redshift $z=0$
in a suite of 30 zoom-in simulations from the Auriga project. 
We compare the results in full magnetohydrodynamics against dark
matter only physics. 
We find a strong influence of baryons in making dark matter halos
rounder at all radii compared to its dark matter only counterparts.
At distances $\lesssim 30$ kpc, rounder and coherent dark matter
distributions correlate with extended massive stellar disks and low gas
densities.  
We measure the alignment between the halo and the stellar disk and
find almost perfect alignment at $0.25R_{200}\sim56$ kpc.   
In some cases the alignment between the two components significantly
changes as a function of radius implying that the halo shape twists;
this effect correlates with extended stellar disks and
low gas densities and is almost absent in the dark matter only
simulations.  
In a comparison against observational constraints  we find that $20\%$
of halos are consistent with observational results derived
from the Pal 5 stream that favor $b/a\approx 1$ and
$c/a>0.77$. Including baryons is a required element to achieve this
level of agreement. In contrast, none of the simulations (dark matter
only and with baryons) can reproduce the constrains derived from the
Saggitarius stream that favor $b/a\approx 0.97$ and $c/a\approx0.44$.
\end{abstract}

@@

Observational constraints of the dark matter halo have been in
conflict during the last decade.
Some studies prefer oblate configurations at small
distances around $\leq 20$ kpc
\citep{LM10,Bovy16,Loebman_et_al._2012,Olling_and_Merrifield_2000,Banerjee_and_Chanda_2011} 
while others tend towards more triaxial and prolate configurations on
the outter distances $\geq 20$ kpc
\citep{Vera-Ciro_and_Helmi_2013,Law_and_Majewski_2009,Banerjee_and_Chanda_2011,Deg_and_Widrow_2013}. 
Some other studies are inclined towards prolate configurations even at the inner
parts of the halo \citep{Bowden_et_al._2016}, and
although it previously seemed that a triaxial DM halo on the
outskirts would be necessary to fully explain the characterization
of the Sagittarius stream \citep{Law_and_Majewski_2009}, recent studies
questioned this claim by reporting inconsistencies with narrow stellar
streams \citet{Pearson_et_al._2015} or finding that
the relaxation of other constraints may make this claim unnecessary
\citet{Ibata_et_al._2013} axisymmetrical to ensure the stability of
a hydrodynamical disk embeded in a static DM halo.

In this paper we address the question of measuring the shape of the DM
halo in MW type galaxies from state-of-the-art simulations from the
Auriga project.
These simulations have large enough numerical resolution, and explicit
cosmological context and an appropriate feedback physics to produce
realistic MW disks.
This paper is structured as follows. 
In Section \ref{sec:numerical} we present the most relevant details of
the simulations, in Section \ref{sec:method} we present the method
that we use to measure the DM halo shape. 
In Section \ref{sec:results} we present our results focusing on the
radial shape trends at $z=0$ and the alignments with the stellar disk.
We discuss our results in Section \ref{sec:discussion} to place them
into the context of other numerical work, explore correlations of the
shape with baryonic properties in the disk and finally make a direct
comparison against observational constraints for the MW's dark matter halo shape.
We finalize with our conclusions in Section \ref{sec:conclusions}.

@@

%\begin{figure*}
%\begin{center}
%\includegraphics[width=0.35\textwidth]{Z_Triax_level3_set_A_DM.pdf}
%\includegraphics[width=0.35\textwidth]{Z_Triax_level3_set_A_MHD.pdf}
%\end{center}
%\caption{Axial ratios as a function of time (colour circles) and
%  radius (dashed lines). 
%  The left/right columns correspond to DMO/MHD simulations, respectively. 
%  The colour indicates the redshift at which the shape was measured
%  (always at $R_{200}$ at that time.)
%  The radial shapes goes from $R_{200}$ down to $\sim 1$ kpc.
%  These three halos correspond to a first half the sample for the \texttt{Level3}
%  haloes. 
%}
%\label{fig:triaxial_history_A}
%\end{figure*}

%\begin{figure*}
%\begin{center}
%\includegraphics[width=0.35\textwidth]{Z_Triax_level3_set_B_DM.pdf}
%\includegraphics[width=0.35\textwidth]{Z_Triax_level3_set_B_MHD.pdf}
%\end{center}
%\caption{Same as Figure \ref{fig:triaxial_history_A}.
%  These three halos correspond the second half of the sample for the
%  \texttt{Level3} haloes. 
%}
%\label{fig:triaxial_history_B}
%\end{figure*}

>>>>

% Abstract of the paper
\begin{abstract}
We present shape measurements of dark matter halos at redshift $z=0$
in a suite of 30 zoom-in simulations from the Auriga project. 
We compare the results in full magnetohydrodynamics against dark
matter only physics. 
We find a strong influence of baryons in making dark matter halos
rounder at all radii compared to its dark matter only counterparts.
At distances $\lesssim 30$ kpc, rounder and coherent dark matter
distributions correlate with extended massive stellar disks and low gas
densities.  
We measure the alignment between the halo and the stellar disk and
find almost perfect alignment at $0.25R_{200}\sim56$ kpc.   
In some cases the alignment between the two components significantly
changes as a function of radius implying that the halo shape twists;
this effect correlates with extended stellar disks and
low gas densities and is almost absent in the dark matter only
simulations.  
In a comparison against observational constraints  we find that $20\%$
of halos are consistent with observational results derived
from the Pal 5 stream that favor an almost spherical shape.
Including baryons is a required element to achieve this
level of agreement. In contrast, none of the simulations (dark matter
only nor with baryons) match the constrains derived from the
Sagittarius stream favoring an oblate dark matter halo.
\end{abstract}

@@

Observational constraints of the dark matter halo have been in
tension during the last decade. 
One can find studies favoring prolate
\citep{Banerjee_and_Chanda_2011,Bowden_et_al._2016},  oblate
\citep{LM10,Deg_and_Widrow_2013,Vera-Ciro_and_Helmi_2013} and
spherical configurations \citep{Bovy16}.  
Constraints from modeling of stellar streams discard the prolate
configuration \citep{Pearson_et_al._2015} although some other studies
still doubt that stellar streams can be used to constrain the halo
shape once other assumptions, such as the density profile, are relaxed
\citep{Ibata_et_al._2013}.

In this paper we contribute to the debate by reporting measurements of
the DM halo shape in MW type galaxies from state-of-the-art simulations from the
Auriga project.
These simulations have large enough numerical resolution, and explicit
cosmological context and an appropriate feedback physics to produce
realistic MW disks.
This paper is structured as follows. 
In Section \ref{sec:numerical} we present the most relevant details of
the simulations, in Section \ref{sec:method} we present the method
that we use to measure the DM halo shape. 
In Section \ref{sec:results} we present our results focusing on the
radial shape trends at $z=0$ and the alignments with the stellar disk.
We discuss our results in Section \ref{sec:discussion} to place them
into the context of other numerical work, explore correlations of the
shape with baryonic properties in the disk and finally make a direct
comparison against observational constraints for the MW's dark matter halo shape.
We finalize with our conclusions in Section \ref{sec:conclusions}.

@@
